BOLD Parameter Estimation Using Particle Filters

### 2010-08-06 doc/thesis/conclusion.tex
fixed up results.
added parameter map and parameter hist
changes in introduction primarily
some changes to the abstract
fixed some captions, and updated conclusion. Various cleanup in methods.

\chapter{Conclusion}
\label{sec:Conclusion}
In this paper I have proposed and demonstrated the use of a particle filter
for the estimation of BOLD model parameters. In doing so, it has been possible
to estimate the BOLD time-series and judge the quality of the estimated time-
series with respect to the true BOLD signal. As result, it is possible to use
the particle filter to localize areas where a known stimuli most directly drives
neural activation. In the future it will be possible to use the estimated states
($v,q,s,f$) to drive other models and learn more about how regions of the brain
interact.

@@

>>>>

\chapter{Conclusion}
\label{sec:Conclusion}
This work has proposed and demonstrated the use of a particle filter
for the estimation of BOLD model parameters. Because many of the parameters
were under constrained from the point of view of the BOLD response, a
full posterior estimate was a more logical approach to parameter estimation
than typical point estimates of parameters. The particle filter also converged
quickly to correct output estimates, and thus provides a framework
for future real-time FMRI experiments. Final BOLD estimates were also comparable
to the results of SPM.

* Overview of results
** Parameters under-constrained
** Output estimates good

* Pros/Cons
** Pro
*** Non-parametric
*** Real time
*** Full Posterior
*** More intuitive
** Cons
*** Computation longer than SPM
*** Harder to interpret

* Enhancements/Dehancements 
** flatten prior
** \% difference using spline?
** "smart knots"
** DC gain as a parameter
** linearizing

* Future works
** Extensive test of quality of volterra kernel estimate from friston
** Better de-trending
** Automatic estimation of measurement error.

As result, it is possible to use
the particle filter to localize areas where a known stimuli most directly drives
neural activation. In the future it will be possible to use the estimated states
($v,q,s,f$) to drive other models and learn more about how regions of the brain
interact.

@@

Automatic detection of the noise level in the signal, to get a decent wieghting
function. Large scale activation to get more parameter estimates. 
Viscoelastic effects from \cite{Buston2004}, discussed in \cite{Deneux2006}.

### 2010-08-06 doc/thesis/methods.tex
fixed some captions, and updated conclusion. Various cleanup in methods.

\chapter{Methods}
\label{sec:Methods}
Although the particle filter I used for this work is a standard Regularized
Particle filter, as described in \cite{Arulampalam2002a}, optimizing the 
particle filter for use with FMRI data is non-trivial.

\section{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied small regions; and most research with
reasonable speed (including \cite{Friston2002b}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to an extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. Consequently for this work,
I wanted to be conservative; which meant going with the accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

>>>>

\chapter{Methods}
\label{sec:Methods}
Although the particle filter  is a standard Regularized
Particle filter, as described in \cite{Arulampalam2002a}, optimizing the 
particle filter for use with FMRI data is non-trivial.

\section{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied regions deemed active by the General 
Linear Model; and most other research (including \cite{Friston2002b}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to a extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. A very wide prior will require
more particles to be sufficiently dense, a very thin prior may miss
the true parameters. Consequently, for this work it was natural
to use priors that will give results consistent with similar tests. accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

### 2010-08-06 doc/thesis/resultsB.tex
fixed some captions, and updated conclusion. Various cleanup in methods.

\begin{figure}[H]
\subfigure[]{\label{fig:hm_spm} \includegraphics[scale=.85]{images/spm_hm}}
\subfigure[]{\label{fig:hm_canon_spm_x} \includegraphics[scale=3]{images/spm_hm_x}}
\subfigure[]{\label{fig:hm_canon_spm_y} \includegraphics[scale=3]{images/spm_hm_y}}
\subfigure[]{\label{fig:hm_canon_spm_z} \includegraphics[scale=3]{images/spm_hm_z}}
\subfigure{\label{fig:scale_spm} \includegraphics[scale=.5]{images/scale1}}
\caption{Sagittal, coronal and axial slices of SPM results (\autoref{fig:hm_canon_spm_x} \autoref{fig:hm_canon_spm_y} 
         \autoref{fig:hm_canon_spm_x}), as well as a series of axial slices, \autoref{fig:hm_spm}. Units
         of activation are in Student's T-scores. Higher indicates higher assurance that the signal cannot
         have occurred through noise alone.}
\label{fig:hm_canon_spm}
\end{figure}

\begin{figure}[H]
\subfigure[]{\label{fig:hm_pfilter} \includegraphics[scale=.85]{images/pfilter_hm}}
\subfigure[]{\label{fig:hm_canon_pfilter_x} \includegraphics[scale=3]{images/pfilter_hm_x}}
\subfigure[]{\label{fig:hm_canon_pfilter_y} \includegraphics[scale=3]{images/pfilter_hm_y}}
\subfigure[]{\label{fig:hm_canon_pfilter_z} \includegraphics[scale=3]{images/pfilter_hm_z}}
\subfigure{\label{fig:scale_pfilter} \includegraphics[scale=.5]{images/scale2}}
\caption{Sagittal, coronal and axial (\autoref{fig:hm_canon_pfilter_x} \autoref{fig:hm_canon_pfilter_y} 
         \autoref{fig:hm_canon_pfilter_x}), as well as a series of axial slices, \autoref{fig:hm_pfilter}. 
         Units of match is normalized residual. The lowest (best) levels were $.7$.
         The highest error shown is $1$.}
\label{fig:hm_canon_pfilter}
\end{figure}

@@

\begin{figure}[H]
\subfigure[]{\label{fig:hm_mi} \includegraphics[scale=.85]{images/mi_hm}}
\subfigure[]{\label{fig:hm_canon_mi_x} \includegraphics[scale=3]{images/mi_hm_x}}
\subfigure[]{\label{fig:hm_canon_mi_y} \includegraphics[scale=3]{images/mi_hm_y}}
\subfigure[]{\label{fig:hm_canon_mi_z} \includegraphics[scale=3]{images/mi_hm_z}}
\subfigure{\label{fig:scale_mi} \includegraphics[scale=.5]{images/scale4}}
\caption{Sagittal, coronal and axial (\autoref{fig:hm_canon_pfilter_x} \autoref{fig:hm_canon_pfilter_y} 
         \autoref{fig:hm_canon_pfilter_x}), as well as a series of axial slices, \autoref{fig:hm_pfilter}. 
         Units of match is bits (standard for base-2 Mutual Information). The highest (best) levels are 
         $.42$. The worst shown is $.1$.}
\label{fig:hm_canon_pfilter_mi}
\end{figure}

>>>>

\begin{figure}[H]
\subfigure[]{\label{fig:hm_spm} \includegraphics[scale=.85]{images/spm_hm}}
\subfigure[]{\label{fig:hm_canon_spm_x} \includegraphics[scale=3]{images/spm_hm_x}}
\subfigure[]{\label{fig:hm_canon_spm_y} \includegraphics[scale=3]{images/spm_hm_y}}
\subfigure[]{\label{fig:hm_canon_spm_z} \includegraphics[scale=3]{images/spm_hm_z}}
\subfigure{\label{fig:scale_spm} \includegraphics[scale=.5]{images/scale1}}
\caption{SPM results. Units of activation are in Student's T-scores; higher indicates higher 
        assurance that the signal cannot have occurred through noise alone. 
        Sagittal, coronal and axial slices  are 
        \autoref{fig:hm_canon_spm_x}, \autoref{fig:hm_canon_spm_y}, and
         \autoref{fig:hm_canon_spm_z}, respectively. A series of axial slices are
         shown in \autoref{fig:hm_spm}. }
\label{fig:hm_canon_spm}
\end{figure}

\begin{figure}[H]
\subfigure[]{\label{fig:hm_pfilter} \includegraphics[scale=.85]{images/pfilter_hm}}
\subfigure[]{\label{fig:hm_canon_pfilter_x} \includegraphics[scale=3]{images/pfilter_hm_x}}
\subfigure[]{\label{fig:hm_canon_pfilter_y} \includegraphics[scale=3]{images/pfilter_hm_y}}
\subfigure[]{\label{fig:hm_canon_pfilter_z} \includegraphics[scale=3]{images/pfilter_hm_z}}
\subfigure{\label{fig:scale_pfilter} \includegraphics[scale=.5]{images/scale2}}
\caption{Particle Filter results measured in normalized $\sqrt{MSE}$. 
        Units of match is normalized residual where the  the lowest (best) levels shown are
        $.7$ and the highest error shown (threshold) is $1$.
        Sagittal, coronal and axial slices are shwon in \autoref{fig:hm_canon_pfilter_x},
        \autoref{fig:hm_canon_pfilter_y}, and 
         \autoref{fig:hm_canon_pfilter_z}, respectively. A series of axial slices are shown in
         \autoref{fig:hm_pfilter}.  }
\label{fig:hm_canon_pfilter}
\end{figure}

@@

\begin{figure}[H]
\subfigure[]{\label{fig:hm_mi} \includegraphics[scale=.85]{images/mi_hm}}
\subfigure[]{\label{fig:hm_canon_mi_x} \includegraphics[scale=3]{images/mi_hm_x}}
\subfigure[]{\label{fig:hm_canon_mi_y} \includegraphics[scale=3]{images/mi_hm_y}}
\subfigure[]{\label{fig:hm_canon_mi_z} \includegraphics[scale=3]{images/mi_hm_z}}
\subfigure{\label{fig:scale_mi} \includegraphics[scale=.5]{images/scale4}}
\caption{Particle Filter results measured in mutual information. 
         Units of match is bits (standard for base-2 Mutual Information). The highest (best) levels are 
         $.42$ and the worst shown (threshold) is $.1$.
        Sagittal, coronal and axial slices are shwon in \autoref{fig:hm_canon_mi_x},
        \autoref{fig:hm_canon_mi_y}, and \autoref{fig:hm_canon_mi_z}, respectively. 
        A series of axial slices are shown in \autoref{fig:hm_mi}.  }
\label{fig:hm_canon_pfilter_mi}
\end{figure}
